Use of general purpose graphics processing units with MODFLOW

### 2012-10-24 GW20120518-0099-r02.tex
revision 02

%%% SETTINGS
\geometry{letterpaper}
\geometry{margin=1in}
\doublespacing
\linenumbers
\allsectionsfont{\sffamily\mdseries\upshape} 
\graphicspath{{./Figures/}}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0pt}
\lhead{}\chead{Hughes and White -- MODFLOW UPCG GPGPU Solver}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

@@

\begin{document}
\maketitle
\begin{abstract}
To evaluate the use of general-purpose graphics processing units (GPGPU) to improve the performance of MODFLOW, an unstructured preconditioned conjugate gradient (UPCG) solver has been developed. The UPCG solver uses a compressed sparse row storage scheme and includes Jacobi, zero fill-in incomplete and modified-incomplete \color{cyan}lower-upper (LU) \color{black}factorization, and generalized least-squares polynomial preconditioners. The UPCG solver also includes options for sequential and parallel solution on the central processing unit (CPU) using OpenMP. For simulations utilizing the GPGPU, all basic linear algebra operations are performed on the GPGPU; memory copies between the central processing unit CPU and GPCPU occur prior to the first iteration of the UPCG solver and after satisfying head and flow criteria or exceeding a maximum number of iterations. The efficiency of the UPCG solver for GPGPU and CPU solutions is benchmarked using simulations of a synthetic, heterogeneous  unconfined aquifer with tens of thousands to millions of active grid cells. Testing indicates GPGPU speedups on the order of 2-8, relative to the standard MODFLOW \color{cyan}preconditioned conjugate gradient (PCG) \color{black}solver, can be achieved when (1) memory copies between the CPU and GPGPU are optimized, (2) the percentage of time performing memory copies between the CPU and GPGPU is small relative to the calculation time, (3) high-performance GPGPU cards are utilized, and (4) CPU-GPGPU combinations are used to execute sequential operations that are difficult to parallelize. Furthermore, UPCG solver testing indicates GPGPU speedups exceed parallel CPU speedups achieved using OpenMP on multi-core CPUs for preconditioners that can be \color{blue}easily \color{black}parallelized.
\end{abstract}

@@

MODFLOW (Harbaugh 2005) is a finite-difference groundwater flow model that has been effectively applied to \color{blue}numerous \color{black}two- and three-dimensional problems. Because MODFLOW uses a cell-centered finite-difference (CCFD) approximation and a rectilinear grid, model runtimes are often increased by factors of two, four, and eight with increased vertical, horizontal, or combined vertical and horizontal discretization, respectively. \color{blue}Increased model runtimes are typically a direct result of an increase in the number of simultaneous equations solved for each cell but can also be affected (typically to a lesser degree) by changes in the coefficient matrix. \color{black}To date, larger model sizes have been accommodated through a combination of faster CPUs and better linear solvers. The original version of MODFLOW (McDonald and Harbaugh 1988) included 1) a strongly implicit procedure (SIP) linear solver and 2) a slice-successive overrelaxation (\color{cyan}SSOR\color{black}) linear solver. These original linear solvers were supplemented with the preconditioned conjugate gradient (PCG) linear solver (Hill 1990), \color{cyan}the alternating diagonal ordering (DE4) direct solver (Harbaugh 1995), \color{black}the link-algebraic multi-grid (LMG) linear solver (Mehl and Hill 2001), the geometric multigrid (GMG) linear solver (Wilson and Naff 2004), and the  preconditioned conjugate gradient with improved nonlinear control (PCGN) linear solver (Naff and Banta 2008).

Parallelization of linear solvers is another approach for accommodating larger model sizes. \color{blue}Ideal parallelization would result in a linear speedup  (CPU runtime / parallel runtime) as additional CPUs and/or computers were applied to any given problem. For example, if ideal parallelization were achieved and 2, 3, 4, and 5 CPUs were applied to a problem, the speedup would be 2, 3, 4, and 5, respectively. However, ideal parallelization is rarely, if ever, achieved because of communication overhead, which may be a result of hardware limitations and/or use of algorithms that are difficult to parallelize. \color{cyan}In this study, speedup is calculated using the amount of time spent in solver routines not the total runtime (which includes the time spent in other non-solver routines).\color{black}

\color{black}Dong and Li (2009) parallelized the MODFLOW PCG solver (Hill 1990) using the open multiprocessing (OpenMP) programming paradigm (OpenMP Architecture Review Board, 2005) and observed speedups as large as 5 on machines with multi-core CPUs \color{cyan}using 8 threads\color{black}. Naff (2008) developed a parallelized linear solver for MODFLOW using the Message Passing Interface (MPI) standard and observed a speedup of 7 \color{cyan}using 32 CPUs\color{black}. In addition to OpenMP and MPI, general-purpose graphics processing units (GPGPUs) also \color{blue}present \color{black}an option for linear solver parallelization. In comparison to CPUs, GPGPUs \color{blue}typically have more cores (\textit{e.g.}, 448 cores in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs), have higher bandwidth, and therefore, are capable of executing more floating point operations per second. \color{black}Use of GPGPUs to parallelize linear solvers for MODFLOW is also attractive because of the availability of the Compute Unified Device Architecture (CUDA) application programming interface (API) for NVIDIA GPGPUs, which supports development of parallel GPGPU code in C/C++ or Python, Fortran, Java, and MATLAB using wrappers around the C/C++ implementation. For example, Soni \textit{et al.} (2012) used the CUDA language to develop a GPGPU code to solve computational fluid dynamics problems. For solution of the groundwater flow equation, Ji \textit{et al.} (\color{cyan}2010 and \color{black}2012) developed a parallel GPGPU version of MODFLOW-2000 with a Jacobi preconditioner and observed speedups ranging from 1.8 to 3.7.

In this paper, we evaluate speedups resulting from GPGPU parallelization of the unstructured preconditioned conjugate gradient (UPCG) linear solver for a unconfined groundwater flow problem. We also evaluate the performance of the UPCG linear solver using Jacobi, zero fill-in modified incomplete  \color{cyan}lower-upper (MILU0) factorization \color{black}, and generalized least-squares polynomial (GLSPOLY) preconditioners on the CPU and GPGPU. And finally, we evaluate the performance of the UPCG solver on the GPGPU to parallel CPU simulations using the OpenMP parallel programming paradigm.

@@

For GPGPU code implementations, the concept is similar to OpenMP.  The main code is a sequential host code, which forks the data into many threads at the invocation of a CUDA \color{blue}functions, called kernels (\textbf{Box 1})\color{black}.  \color{blue}However, as opposed to the few threads spawned by OpenMP, GPGPUs have the potential to spawn thousands to tens of thousands of threads to execute the CUDA kernel function (\textbf{Box 2}). The NVIDIA Tesla C2050 used to evaluate the UPCG solver in this study can execute 21,504 concurrent threads. \color{black} With thread counts on the order of tens of thousands, GPGPUs have great potential for parallelization \color{blue} for MODFLOW and other numerical simulators that use Krylov subspace methods to solve large systems of simultaneous linear equations\color{black}.

@@

In \textbf{algorithm~\ref{pcgalg}}, $\mathbf{r}$ is the \color{cyan}difference between the left and right side (residual) \color{black}of \textbf{equation~\ref{lineareq}} [L$^3$/T], $i$ is the linear iteration index [unitless], $maxinner$ is a user-specified maximum number of iterations to perform [unitless], $\mathbf{M}$ is the preconditioned form of $\mathbf{A}$ [L$^2$/T], $\mathbf{z}$ is the solution resulting from application of $\mathbf{M}$ to $\mathbf{r}$ [L$^3$/T], $\mathbf{p}$ is the orthogonal search direction [L], $\beta$ is a scalar used to determine the next search direction [unitless], $\mathbf{q}$ is the residual change resulting from multiplication of $\mathbf{A}$ and the search direction $\mathbf{p}$ [L$^3$/T], $\alpha$ is the step-size [unitless]. \color{cyan}Readers are referred to Barrett \textit{et al.} (1994) for more information on the PCG method.\color{black}

All linear iteration steps, except for application of the zero fill-in incomplete LU (ILU0) and MILU0 preconditioners in line 3 of \textbf{algorithm~\ref{pcgalg}}, include operations that are independent for each active groundwater cell, indicating a potential for parallelization. In cases where (1) no preconditioner ($\mathbf{z = r}$) or (2) a Jacobi preconditioner ($\mathbf{M = I D}^{-1}$ -- where $\mathbf{I}$ is the identity matrix and $\mathbf{D}^{-1}$ is a matrix that only contains the inverse of the diagonal elements of $\mathbf{A}$), is selected, application of the preconditioner is highly parallelizable. For most practical problems, non-preconditioned and the Jacobi preconditioners will not significantly reduce the number of linear iterations ($i$) required to achieve convergence (Li and Saad 2010). As a result better preconditioners, such as incomplete factorizations (ILU0 or MILU0) or m-degree polynomial (GLSPOLY) approximations of $\mathbf{A}$, are generally needed. Application of the incomplete factorization preconditioners requires a forward substitution ($\mathbf{Ly = b}$ -- where $\mathbf{L}$ is the lower triangular portion of $\mathbf{A}$) followed by a backward substitution ($\mathbf{Uz = y}$ – where $\mathbf{U}$ is upper triangular portion of $\mathbf{A}$) and is difficult to efficiently parallelize (Barrett \textit{et al.} 1994). Preconditioning using polynomial preconditioners, however, is simple, efficient, and easy to parallelize \color{cyan}because polynomial preconditioner application only requires a limited number of operations (Liang \textit{et al.} 2002; Saad 2003); the operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector multiplied by a constant to another vector (\texttt{axpy}).\color{black}

\subsection*{Numerical Implementation}
The \color{cyan}unstructured preconditioned conjugate gradient (UPCG) \color{black}linear solver was coded to allow solution of the groundwater flow equation on either the CPU, GPGPU, or combination of CPU and GPGPU using double precision operations. Execution of the UPCG solver on the CPU, GPGPU, or combination of CPU and GPGPU is specified by the user in the UPCG input file.

Although the connectivity of the finite-difference discretization used in MODFLOW uses a fixed 5- and 7-point stencil in two- and three-dimensions, respectively, the PCG method in \textbf{algorithm~\ref{pcgalg}} was coded using an unstructured compressed sparse row (CSR) storage format  (Barrett \textit{et al.} 1994) with the diagonal element in the first position in each row. This storage format is also supported by the \color{blue}CUDA-enabled implementation of the BLAS library (CUBLAS -- NVIDIA 2012). BLAS is a library of Basic Linear Algebra Subprograms (Lawson \textit{et al.} 1979)\color{black}. The UPCG linear solver includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners (\color{cyan}Saad \color{black}2003).

@@

For parallel CPU simulations, OpenMP directives were added to all matrix-vector products and vector operations. The OpenMP matrix-vector product was constructed using a level-2 BLAS \color{cyan}general matrix-vector (GEMV) product \color{black}approach because of the slight increase in performance of this approach over dot-product and \texttt{axpy} approaches for calculating matrix vector products (Petersen and Arbenz 2004). OpenMP directives \texttt{NUM\_THREADS} and \texttt{SCHEDULE STATIC} were used in all OpenMP routines to allow the user to control the number of OpenMP threads and reduce OpenMP scheduling/synchronization overhead, respectively. The number of OpenMP threads used for matrix-vector products and vector operations is specified separately in the UPCG input file. For sequential CPU simulations, separate matrix-vector product and vector operation routines that exclude OpenMP directives are used to eliminate all OpenMP overhead.

Infinity norms ($\left \|  x \right \|_{\infty}$) are used to evaluate convergence of the UPCG linear solver with respect to the change in $\mathbf{h}$ (HCLOSE) and $\mathbf{r}$ (RCLOSE). In MODFLOW, non-linearities are resolved using Picard iteration. Picard iteration is implemented such that the $\mathbf{A}$ matrix is formulated using the latest estimate of $\mathbf{h}$, then $\mathbf{h}$ and $\mathbf{r}$ are updated using a linear solver. This process is continued until the maximum change in $h$ and $r$ in any model cell is less than HCLOSE and RCLOSE, respectively, on the first iteration ($i$=1) of the linear solver or the maximum number of Picard (outer) iterations is exceeded; \color{cyan}the UPCG solver approach for evaluating convergence is identical to the standard MODFLOW PCG solver (Hill 1990).\color{black}

\section* {\bf Test Cases}
The performance of the GPGPU implementation of the UPCG linear solver was evaluated on a NVIDIA Tesla C2050 GPGPU with 3 GB of RAM and capable of 515 GFLOP/sec of double precision processing performance (NVIDIA compute capability 2.0). The GPGPU was mounted in a 16-pin PCI local bus on a single quad core Intel \color{cyan}Xeon \color{black}3GHz CPU, capable of executing 4 threads per CPU, with 6 GB of RAM and running a 64-bit version of the Windows 7 Enterprise OS. For comparison with GPGPU results, \color{cyan}serial and \color{black}parallel CPU simulations were executed on dual 4-core Intel \color{cyan}Xeon \color{black}2.4 GHz CPUs, capable of executing 8 parallel threads, with 16 GB of RAM running a 64-bit version of the Windows Server 2008 R2 Standard OS (Service Pack 1).

\color{blue}To evaluate the performance of the GPGPU implementation of the UPCG linear solver a  synthetic test problem with aquifer properties and boundary conditions that result in equivalent problems and scalable runtimes when discretized using different horizontal and vertical cell sizes. The test problem represents an unconfined aquifer to increase the difficulty of the problem by making it non-linear because of the dependence of the $\mathbf{A}$ matrix in \textbf{equation~\ref{lineareq}} on the solution of $\mathbf{h}$. Furthermore, the test problems used realistic initial conditions sufficiently far from final results to start with an incorrect $\mathbf{A}$ matrix to maximize (1) head changes during solution of the problem and (2) the number of Picard iterations required to achieve a solution.\color{black}

\subsection*{Problem Description}
GPGPU and parallel CPU results were evaluated using a number of horizontal and vertical model discretizations of a \color{cyan}1,000 m $\times$ 1,000 m \color{black} heterogeneous unconfined aquifer. The top and bottom of the aquifer are flat and specified to have elevations of 10 and -30 m, respectively. Models with horizontal discretizations of 200 $\times$ 200, 500 $\times$ 500, 1,000 $\times$ 1,000, 2,000 $\times$ 2,000, and 4,000 $\times$ 4,000 cells per layer were evaluated. Single layer models were evaluated for all horizontal discretizations and multi-layer models were evaluated for select horizontal discretizations to determine the effect that a 7-point stencil, and the resulting additional non-zero entries for each cell, has on GPGPU and parallel CPU execution time. Two- and three-layer models were evaluated for all horizontal discretizations except 4,000 $\times$ 4,000 cells per layer. Ten-layer models were also evaluated for horizontal discretizations ranging from 200 $\times$ 200 to 1,000 $\times$ 1,000 cells per layer. Additional horizontal and vertical discretizations were not evaluated because of GPGPU memory requirements for larger problems. For multi-layer models, a constant layer thickness was used for all but the first layer and was calculated by dividing 30 m by the number of layers. \color{blue}Layer 1 also includes the thickness of the aquifer above the arbitrary elevation of 0 m and therefore is 10 m thicker than layer 2.\color{black}

@@

Head-dependent (general head) boundaries were specified on the left and right sides \color{cyan} in all layers \color{black} and no-flow boundaries at the top and bottom. Head-dependent boundaries are assumed to be located a half model cell away from the edge of the model domain and conductance values were calculated using the average hydraulic conductivity of 4.81 m/d, vertical saturated cell areas, and half the model cell dimensions in the x-direction; head-dependent boundary head values were calculated using the hydraulic gradient over the model domain \color{blue}$\left [ 10. / 1000. = 0.01 \text{ m/m} \right]$ \color{black}and assumed heads at the left and right sides of the model domain of 10.0 and 0.0 m, respectively. For the one layer 1,000 $\times$ 1,000 cell model, conductances and head values of 214.1447 m$^{2}$/d, 160.6085 m$^{2}$/d, 10.025 m, and -0.025 m were specified for the left and right side of the model domain, respectively. The head-dependent boundaries cause an ambient groundwater flow from left to right. Four pumping wells, located in the four cells in the center of the model domain, pump at a constant rate with a total groundwater withdrawal rate of 1,000 m$^3$/d. For multi-layer simulations, the pumping wells withdraw water from the lower model layer. There is no recharge or evapotranspiration; all groundwater flow to the wells is provided by the head-dependent boundaries \color{cyan}and represents 46\% of the flow entering the model\color{black}. Boundary conditions are shown graphically in \textbf{figure~\ref{FigModelDomain}}.

@@

All of the models simulated steady-state conditions. As a result, specific storage and specific yield values were not specified for the aquifer. Initial heads were specified to be 0.0 m throughout the model domain, sufficiently different from final steady-state heads. Simulated heads \color{cyan} that satisfy specified convergence criteria \color{black} for the one layer 1,000 $\times$ 1,000 cell model are shown in \textbf{figure~\ref{FigModelDomain}} and show the effect that groundwater withdrawals and heterogeneous hydraulic conductivity have on water levels. \color{cyan}Water-level distributions for all converged simulations, including simulations using different (1) solvers, (2) preconditioners, and/or (3) hardware, are identical within the tolerance of $\pm$HCLOSE m.\color{black}

\subsection*{Results}
\color{blue}The number of Picard and inner iterations required to solve the test problems on the CPU are summarized on \textbf{table~\ref{TableCPUITSummary}}; the number of Picard and inner iterations required to solve the test problems on the GPGPU and hybrid CPU/GPGPU approaches were identical to the number required on the CPU. The number of Picard and inner iterations required by the MILU0 preconditioner is comparable to the standard MODFLOW PCG solver with the modified incomplete Cholesky (MIC) preconditioner (Hill 1990). The number of Picard iterations is also comparable between the MILU0 preconditioner, PCG solver with the MIC preconditioner, and the GMG solver with the ILU0 preconditioner; the GMG inner iterations are not comparable because of differences in the inner iteration approaches of both the UPCG and PCG solvers and the GMG solver. The number of Picard and inner iterations required by the GLSPOLY preconditioner is generally greater than the MILU0 preconditioner and the PCG solver with the MIC preconditioner but it is notably better the Jacobi preconditioner. In fact, the maximum number of Picard and inner (linear) iterations were exceeded for the 4,000 $\times$ 4,000 $\times$ 1 cell problem simulated using the Jacobi preconditioner, highlighting that the Jacobi preconditioner is not a robust preconditioner for some problems.\color{black}

>>>>

%%% SETTINGS
\geometry{letterpaper}
\geometry{margin=1in}
\doublespacing
%\linenumbers
\allsectionsfont{\sffamily\mdseries\upshape} 
\graphicspath{{./Figures/}}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0pt}
%\lhead{}\chead{Hughes and White -- MODFLOW UPCG GPGPU Solver}\rhead{}
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

@@

\begin{document}
\maketitle
\begin{abstract}
To evaluate the use of general-purpose graphics processing units (GPGPU) to improve the performance of MODFLOW, an unstructured preconditioned conjugate gradient (UPCG) solver has been developed. The UPCG solver uses a compressed sparse row storage scheme and includes Jacobi, zero fill-in incomplete and modified-incomplete lower-upper (LU) factorization, and generalized least-squares polynomial preconditioners. The UPCG solver also includes options for sequential and parallel solution on the central processing unit (CPU) using OpenMP. For simulations utilizing the GPGPU, all basic linear algebra operations are performed on the GPGPU; memory copies between the central processing unit CPU and GPCPU occur prior to the first iteration of the UPCG solver and after satisfying head and flow criteria or exceeding a maximum number of iterations. The efficiency of the UPCG solver for GPGPU and CPU solutions is benchmarked using simulations of a synthetic, heterogeneous  unconfined aquifer with tens of thousands to millions of active grid cells. Testing indicates GPGPU speedups on the order of 2-8, relative to the standard MODFLOW preconditioned conjugate gradient (PCG) solver, can be achieved when (1) memory copies between the CPU and GPGPU are optimized, (2) the percentage of time performing memory copies between the CPU and GPGPU is small relative to the calculation time, (3) high-performance GPGPU cards are utilized, and (4) CPU-GPGPU combinations are used to execute sequential operations that are difficult to parallelize. Furthermore, UPCG solver testing indicates GPGPU speedups exceed parallel CPU speedups achieved using OpenMP on multi-core CPUs for preconditioners that can be easily parallelized.
\end{abstract}

@@

MODFLOW (Harbaugh 2005) is a finite-difference groundwater flow model that has been effectively applied to numerous two- and three-dimensional problems. Because MODFLOW uses a cell-centered finite-difference (CCFD) approximation and a rectilinear grid, model runtimes are often increased by factors of two, four, and eight with increased vertical, horizontal, or combined vertical and horizontal discretization, respectively. Increased model runtimes are typically a direct result of an increase in the number of simultaneous equations solved for each cell but can also be affected (typically to a lesser degree) by changes in the coefficient matrix. To date, larger model sizes have been accommodated through a combination of faster CPUs and better linear solvers. The original version of MODFLOW (McDonald and Harbaugh 1988) included 1) a strongly implicit procedure (SIP) linear solver and 2) a slice-successive overrelaxation (SSOR) linear solver. These original linear solvers were supplemented with the preconditioned conjugate gradient (PCG) linear solver (Hill 1990), the alternating diagonal ordering (DE4) direct solver (Harbaugh 1995), the link-algebraic multi-grid (LMG) linear solver (Mehl and Hill 2001), the geometric multigrid (GMG) linear solver (Wilson and Naff 2004), and the  preconditioned conjugate gradient with improved nonlinear control (PCGN) linear solver (Naff and Banta 2008).

Parallelization of linear solvers is another approach for accommodating larger model sizes. Ideal parallelization would result in a linear speedup  (CPU runtime / parallel runtime) as additional CPUs and/or computers were applied to any given problem. For example, if ideal parallelization were achieved and 2, 3, 4, and 5 CPUs were applied to a problem, the speedup would be 2, 3, 4, and 5, respectively. However, ideal parallelization is rarely, if ever, achieved because of communication overhead, which may be a result of hardware limitations and/or use of algorithms that are difficult to parallelize. In this study, speedup is calculated using the amount of time spent in solver routines not the total runtime (which includes the time spent in other non-solver routines).

Dong and Li (2009) parallelized the MODFLOW PCG solver (Hill 1990) using the open multiprocessing (OpenMP) programming paradigm (OpenMP Architecture Review Board, 2005) and observed speedups as large as 5 on machines with multi-core CPUs using 8 threads. Naff (2008) developed a parallelized linear solver for MODFLOW using the Message Passing Interface (MPI) standard and observed a speedup of 7 using 32 CPUs. In addition to OpenMP and MPI, general-purpose graphics processing units (GPGPUs) also present an option for linear solver parallelization. In comparison to CPUs, GPGPUs typically have more cores (\textit{e.g.}, 448 cores in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs), have higher bandwidth, and therefore, are capable of executing more floating point operations per second. Use of GPGPUs to parallelize linear solvers for MODFLOW is also attractive because of the availability of the Compute Unified Device Architecture (CUDA) application programming interface (API) for NVIDIA GPGPUs, which supports development of parallel GPGPU code in C/C++ or Python, Fortran, Java, and MATLAB using wrappers around the C/C++ implementation. For example, Soni \textit{et al.} (2012) used the CUDA language to develop a GPGPU code to solve computational fluid dynamics problems. For solution of the groundwater flow equation, Ji \textit{et al.} (2010 and 2012) developed a parallel GPGPU version of MODFLOW-2000 with a Jacobi preconditioner and observed speedups ranging from 1.8 to 3.7.

In this paper, we evaluate speedups resulting from GPGPU parallelization of the unstructured preconditioned conjugate gradient (UPCG) linear solver for a unconfined groundwater flow problem. We also evaluate the performance of the UPCG linear solver using Jacobi, zero fill-in modified incomplete  lower-upper (MILU0) factorization , and generalized least-squares polynomial (GLSPOLY) preconditioners on the CPU and GPGPU. And finally, we evaluate the performance of the UPCG solver on the GPGPU to parallel CPU simulations using the OpenMP parallel programming paradigm.

@@

For GPGPU code implementations, the concept is similar to OpenMP.  The main code is a sequential host code, which forks the data into many threads at the invocation of a CUDA functions, called kernels (\textbf{Box 1}).  However, as opposed to the few threads spawned by OpenMP, GPGPUs have the potential to spawn thousands to tens of thousands of threads to execute the CUDA kernel function (\textbf{Box 2}). The NVIDIA Tesla C2050 used to evaluate the UPCG solver in this study can execute 21,504 concurrent threads.  With thread counts on the order of tens of thousands, GPGPUs have great potential for parallelization  for MODFLOW and other numerical simulators that use Krylov subspace methods to solve large systems of simultaneous linear equations.

@@

In \textbf{algorithm~\ref{pcgalg}}, $\mathbf{r}$ is the difference between the left and right side (residual) of \textbf{equation~\ref{lineareq}} [L$^3$/T], $i$ is the linear iteration index [unitless], $maxinner$ is a user-specified maximum number of iterations to perform [unitless], $\mathbf{M}$ is the preconditioned form of $\mathbf{A}$ [L$^2$/T], $\mathbf{z}$ is the solution resulting from application of $\mathbf{M}$ to $\mathbf{r}$ [L$^3$/T], $\mathbf{p}$ is the orthogonal search direction [L], $\beta$ is a scalar used to determine the next search direction [unitless], $\mathbf{q}$ is the residual change resulting from multiplication of $\mathbf{A}$ and the search direction $\mathbf{p}$ [L$^3$/T], $\alpha$ is the step-size [unitless]. Readers are referred to Barrett \textit{et al.} (1994) for more information on the PCG method.

All linear iteration steps, except for application of the zero fill-in incomplete LU (ILU0) and MILU0 preconditioners in line 3 of \textbf{algorithm~\ref{pcgalg}}, include operations that are independent for each active groundwater cell, indicating a potential for parallelization. In cases where (1) no preconditioner ($\mathbf{z = r}$) or (2) a Jacobi preconditioner ($\mathbf{M = I D}^{-1}$ -- where $\mathbf{I}$ is the identity matrix and $\mathbf{D}^{-1}$ is a matrix that only contains the inverse of the diagonal elements of $\mathbf{A}$), is selected, application of the preconditioner is highly parallelizable. For most practical problems, non-preconditioned and the Jacobi preconditioners will not significantly reduce the number of linear iterations ($i$) required to achieve convergence (Li and Saad 2010). As a result better preconditioners, such as incomplete factorizations (ILU0 or MILU0) or m-degree polynomial (GLSPOLY) approximations of $\mathbf{A}$, are generally needed. Application of the incomplete factorization preconditioners requires a forward substitution ($\mathbf{Ly = b}$ -- where $\mathbf{L}$ is the lower triangular portion of $\mathbf{A}$) followed by a backward substitution ($\mathbf{Uz = y}$ – where $\mathbf{U}$ is upper triangular portion of $\mathbf{A}$) and is difficult to efficiently parallelize (Barrett \textit{et al.} 1994). Preconditioning using polynomial preconditioners, however, is simple, efficient, and easy to parallelize because polynomial preconditioner application only requires a limited number of operations (Liang \textit{et al.} 2002; Saad 2003); the operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector multiplied by a constant to another vector (\texttt{axpy}).

\subsection*{Numerical Implementation}
The unstructured preconditioned conjugate gradient (UPCG) linear solver was coded to allow solution of the groundwater flow equation on either the CPU, GPGPU, or combination of CPU and GPGPU using double precision operations. Execution of the UPCG solver on the CPU, GPGPU, or combination of CPU and GPGPU is specified by the user in the UPCG input file.

Although the connectivity of the finite-difference discretization used in MODFLOW uses a fixed 5- and 7-point stencil in two- and three-dimensions, respectively, the PCG method in \textbf{algorithm~\ref{pcgalg}} was coded using an unstructured compressed sparse row (CSR) storage format  (Barrett \textit{et al.} 1994) with the diagonal element in the first position in each row. This storage format is also supported by the CUDA-enabled implementation of the BLAS library (CUBLAS -- NVIDIA 2012). BLAS is a library of Basic Linear Algebra Subprograms (Lawson \textit{et al.} 1979). The UPCG linear solver includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners (Saad 2003).

@@

For parallel CPU simulations, OpenMP directives were added to all matrix-vector products and vector operations. The OpenMP matrix-vector product was constructed using a level-2 BLAS general matrix-vector (GEMV) product approach because of the slight increase in performance of this approach over dot-product and \texttt{axpy} approaches for calculating matrix vector products (Petersen and Arbenz 2004). OpenMP directives \texttt{NUM\_THREADS} and \texttt{SCHEDULE STATIC} were used in all OpenMP routines to allow the user to control the number of OpenMP threads and reduce OpenMP scheduling/synchronization overhead, respectively. The number of OpenMP threads used for matrix-vector products and vector operations is specified separately in the UPCG input file. For sequential CPU simulations, separate matrix-vector product and vector operation routines that exclude OpenMP directives are used to eliminate all OpenMP overhead.

Infinity norms ($\left \|  x \right \|_{\infty}$) are used to evaluate convergence of the UPCG linear solver with respect to the change in $\mathbf{h}$ (HCLOSE) and $\mathbf{r}$ (RCLOSE). In MODFLOW, non-linearities are resolved using Picard iteration. Picard iteration is implemented such that the $\mathbf{A}$ matrix is formulated using the latest estimate of $\mathbf{h}$, then $\mathbf{h}$ and $\mathbf{r}$ are updated using a linear solver. This process is continued until the maximum change in $h$ and $r$ in any model cell is less than HCLOSE and RCLOSE, respectively, on the first iteration ($i$=1) of the linear solver or the maximum number of Picard (outer) iterations is exceeded; the UPCG solver approach for evaluating convergence is identical to the standard MODFLOW PCG solver (Hill 1990).

\section* {\bf Test Cases}
The performance of the GPGPU implementation of the UPCG linear solver was evaluated on a NVIDIA Tesla C2050 GPGPU with 3 GB of RAM and capable of 515 GFLOP/sec of double precision processing performance (NVIDIA compute capability 2.0). The GPGPU was mounted in a 16-pin PCI local bus on a single quad core Intel Xeon 3GHz CPU, capable of executing 4 threads per CPU, with 6 GB of RAM and running a 64-bit version of the Windows 7 Enterprise OS. For comparison with GPGPU results, serial and parallel CPU simulations were executed on dual 4-core Intel Xeon 2.4 GHz CPUs, capable of executing 8 parallel threads, with 16 GB of RAM running a 64-bit version of the Windows Server 2008 R2 Standard OS (Service Pack 1).

To evaluate the performance of the GPGPU implementation of the UPCG linear solver a  synthetic test problem with aquifer properties and boundary conditions that result in equivalent problems and scalable runtimes when discretized using different horizontal and vertical cell sizes. The test problem represents an unconfined aquifer to increase the difficulty of the problem by making it non-linear because of the dependence of the $\mathbf{A}$ matrix in \textbf{equation~\ref{lineareq}} on the solution of $\mathbf{h}$. Furthermore, the test problems used realistic initial conditions sufficiently far from final results to start with an incorrect $\mathbf{A}$ matrix to maximize (1) head changes during solution of the problem and (2) the number of Picard iterations required to achieve a solution.

\subsection*{Problem Description}
GPGPU and parallel CPU results were evaluated using a number of horizontal and vertical model discretizations of a 1,000 m $\times$ 1,000 m  heterogeneous unconfined aquifer. The top and bottom of the aquifer are flat and specified to have elevations of 10 and -30 m, respectively. Models with horizontal discretizations of 200 $\times$ 200, 500 $\times$ 500, 1,000 $\times$ 1,000, 2,000 $\times$ 2,000, and 4,000 $\times$ 4,000 cells per layer were evaluated. Single layer models were evaluated for all horizontal discretizations and multi-layer models were evaluated for select horizontal discretizations to determine the effect that a 7-point stencil, and the resulting additional non-zero entries for each cell, has on GPGPU and parallel CPU execution time. Two- and three-layer models were evaluated for all horizontal discretizations except 4,000 $\times$ 4,000 cells per layer. Ten-layer models were also evaluated for horizontal discretizations ranging from 200 $\times$ 200 to 1,000 $\times$ 1,000 cells per layer. Additional horizontal and vertical discretizations were not evaluated because of GPGPU memory requirements for larger problems. For multi-layer models, a constant layer thickness was used for all but the first layer and was calculated by dividing 30 m by the number of layers. Layer 1 also includes the thickness of the aquifer above the arbitrary elevation of 0 m and therefore is 10 m thicker than layer 2.

@@

Head-dependent (general head) boundaries were specified on the left and right sides  in all layers  and no-flow boundaries at the top and bottom. Head-dependent boundaries are assumed to be located a half model cell away from the edge of the model domain and conductance values were calculated using the average hydraulic conductivity of 4.81 m/d, vertical saturated cell areas, and half the model cell dimensions in the x-direction; head-dependent boundary head values were calculated using the hydraulic gradient over the model domain $\left [ 10. / 1000. = 0.01 \text{ m/m} \right]$ and assumed heads at the left and right sides of the model domain of 10.0 and 0.0 m, respectively. For the one layer 1,000 $\times$ 1,000 cell model, conductances and head values of 214.1447 m$^{2}$/d, 160.6085 m$^{2}$/d, 10.025 m, and -0.025 m were specified for the left and right side of the model domain, respectively. The head-dependent boundaries cause an ambient groundwater flow from left to right. Four pumping wells, located in the four cells in the center of the model domain, pump at a constant rate with a total groundwater withdrawal rate of 1,000 m$^3$/d. For multi-layer simulations, the pumping wells withdraw water from the lower model layer. There is no recharge or evapotranspiration; all groundwater flow to the wells is provided by the head-dependent boundaries and represents 46\% of the flow entering the model. Boundary conditions are shown graphically in \textbf{figure~\ref{FigModelDomain}}.

@@

All of the models simulated steady-state conditions. As a result, specific storage and specific yield values were not specified for the aquifer. Initial heads were specified to be 0.0 m throughout the model domain, sufficiently different from final steady-state heads. Simulated heads  that satisfy specified convergence criteria  for the one layer 1,000 $\times$ 1,000 cell model are shown in \textbf{figure~\ref{FigModelDomain}} and show the effect that groundwater withdrawals and heterogeneous hydraulic conductivity have on water levels. Water-level distributions for all converged simulations, including simulations using different (1) solvers, (2) preconditioners, and/or (3) hardware, are identical within the tolerance of $\pm$HCLOSE m.

\subsection*{Results}
The number of Picard and inner iterations required to solve the test problems on the CPU are summarized on \textbf{table~\ref{TableCPUITSummary}}; the number of Picard and inner iterations required to solve the test problems on the GPGPU and hybrid CPU/GPGPU approaches were identical to the number required on the CPU. The number of Picard and inner iterations required by the MILU0 preconditioner is comparable to the standard MODFLOW PCG solver with the modified incomplete Cholesky (MIC) preconditioner (Hill 1990). The number of Picard iterations is also comparable between the MILU0 preconditioner, PCG solver with the MIC preconditioner, and the GMG solver with the ILU0 preconditioner; the GMG inner iterations are not comparable because of differences in the inner iteration approaches of both the UPCG and PCG solvers and the GMG solver. The number of Picard and inner iterations required by the GLSPOLY preconditioner is generally greater than the MILU0 preconditioner and the PCG solver with the MIC preconditioner but it is notably better the Jacobi preconditioner. In fact, the maximum number of Picard and inner (linear) iterations were exceeded for the 4,000 $\times$ 4,000 $\times$ 1 cell problem simulated using the Jacobi preconditioner, highlighting that the Jacobi preconditioner is not a robust preconditioner for some problems.
